---
title: "17: Two sample Confidence Intervals"
author: "Prof Amanda Luby"
subtitle: "Stat250 S25"
callout-appearance: minimal
knitr:
    opts_chunk: 
      dev: "ragg_png"
      echo: false
      warning: false
      message: false
format:
  pdf:
    include-in-header: 
       - "../preamble.tex"
    toc: false
    number-sections: true
    colorlinks: true
    geometry:
      - top=1in
      - left=1in
      - right=1in
      - bottom=1in
      - heightrounded
    mainfont: "Linux Libertine"
    sansfont: "Linux Libertine"
    mathfont: "Libertinus Math"
    monofont: "Monaco"
fontsize: 11pt
---

```{r}
library(tidyverse)
library(ggformula)
library(knitr)
library(mosaic)
library(patchwork)
theme_set(theme_minimal())
```

```{r}
bullets <- readxl::read_excel("monson.xlsx", sheet = 2) |>
  group_by(ExamID) |>
  summarize(
    er_bb = mean(`Score` %in% c("Incorrect-actually Match(gt)", "Incorrect-actually Non Match(gt)")),
    inc_bb = mean(Score %in% c("Match(gt) marked Inconclusive", "Non-Matched(gt) marked Inconclusive"))
  )

cc <- readxl::read_excel("monson.xlsx", sheet = 3) |>
  group_by(ExamID) |>
  summarize(
    er_cc = mean(`Score` %in% c("Incorrect-actually Match(gt)", "Incorrect-actually Non Match(gt)")),
    inc_cc = mean(Score %in% c("Match(gt) marked Inconclusive", "Non-Matched(gt) marked Inconclusive"))
  )

set.seed(0508)
firearms <- bullets |>
  left_join(cc, by = "ExamID") |>
  slice_sample(n=40)
```

Goal: develop an **interval estimate** for a difference in population means: $\mu_1 - \mu_2$

**Example:** There are two types of forensic analyses that are typically performed in cases where firearms evidence has been collected: bullets and cartridges. We are interested in whether there is a difference in the average *inconclusive rate* (how often the firearms examiner cannot come to a definitive conclusion) for these two evidence types. Some results from a study are included below:

```{r}
#| fig-width: 3
#| fig-height: 2
#| layout-ncol: 2

firearms0 = firearms
firearms0[1:2, 2:3] = NA

firearms0 |>
  pivot_longer(c(inc_bb, inc_cc)) |>
  ggplot(aes(x = value, fill = name)) + 
  geom_histogram(bins = 15, col = "white") + 
  scale_fill_viridis_d(end = .75, option = "plasma") + 
  facet_wrap(~name) + 
  theme(legend.position = "none")

firearms0 |>
  pivot_longer(c(inc_bb, inc_cc)) |>
  mutate(
    type = ifelse(name == "inc_bb", "bullet", "cartridge")
  ) |>
  group_by(type) |>
  summarize(
    mean = mean(value, na.rm = TRUE),
    sd = sd(value, na.rm = TRUE),
    n = sum(!is.na(value))
  ) |>
  knitr::kable(digits = 3)
```

Let's review our confidence interval toolkit so far:

1.  Construct a bootstrap distribution and find a **percentile-based** confidence interval

2.  Either (a) assume a normal population or (b) use the CLT to construct a **formula t-based** confidence interval

3.  Use the bootstrap to find the distribution of the T statistics to construct a **bootstrap t** confidence interval

### Bootstrap percentile interval {.unnumbered}

```{r}
bullets_inc <- firearms0$inc_bb[!is.na(firearms0$inc_bb)]
cc_inc <- firearms0$inc_cc[!is.na(firearms0$inc_cc)]
```

```{r}
#| echo: true 
x <- bullets_inc
y <- cc_inc

n1 = length(x)
n2 = length(y)
N = 10^4
boot_diff = numeric(N)

for(i in 1:N){
  x_boot <- sample(x, size = n1, replace = TRUE)
  y_boot <- sample(y, size = n2, replace = TRUE)
  boot_diff[i] <- mean(x_boot, na.rm = TRUE) - mean(y_boot, na.rm = TRUE)
}

quantile(boot_diff, probs = c(.025, .975))
```

### CLT-based t confidence interval {.unnumbered}

```{r}
#| echo: true 

qt(.975, df = 29)
```

\vspace{1in}

```{r}
#| echo: true 

t.test(bullets_inc, cc_inc, conf.level = .95)$conf
```


### Bootstrap t confidence interval {.unnumbered}


```{r}
#| echo: true 
n1 = length(bullets_inc)
n2 = length(cc_inc)
N = 10^4
boot_tstar <- numeric(N)
for(i in 1:N){
  x_boot <- sample(x, size = n1, replace = TRUE)
  y_boot <- sample(y, size = n2, replace = TRUE)
  boot_tstar[i] <- ((mean(x_boot) - mean(y_boot)) - (mean(x) - mean(y)))/sqrt(var(x_boot)/n1 + var(y_boot)/n2)
}

quantile(boot_tstar, probs = c(.025, .975))
(mean(x)- mean(y)) - quantile(boot_tstar, c(.975, .025))*sqrt(var(x)/n1 + var(y)/n2)
```

::: callout-warning
## Watch out!

Independent vs Paired
:::


**Example:** Oops! I forgot to tell you that this study actually had the same examiners in both samples. (There are just two missing values in `bullets_inc`). 

\vspace{2in}

```{r}
#| echo: true


x <- bullets_inc - cc_inc[-c(1,2)]
n = length(x)
N = 10^4
boot_diff = numeric(N)

for(i in 1:N){
  x_boot <- sample(x, size = n, replace = TRUE)
  boot_diff[i] <- mean(x_boot, na.rm = TRUE)
}

quantile(boot_diff, probs = c(.025, .975))
```


```{r}
#| echo: true 


t.test(bullets_inc, cc_inc[-c(1,2)], paired = TRUE)$conf
```

