---
title: "10: Asymptotic Properties"
format: 
  html:
    code-tools: true
---

```{r setup}
#| include: false

knitr::opts_chunk$set(echo = TRUE,
                  message = FALSE,
                  warning = FALSE)
```

```{r}
library(tidyverse)
library(patchwork)
```

# Example from Notes09: Exponential distribution

## Example 1:

Let $Y_1, ..., Y_n$ be $n$ Exp($\lambda$) random variables Let $\widehat\lambda = \frac{n}{\sum Y_i}$. How does Var($\widehat\lambda$) compare with the CRLB?

```{r}
lambda <- .5
n <- 20
n_sims <- 10000
lambda_hat <- numeric(n_sims)

for(i in 1:n_sims){
  sample <- rexp(n, rate = lambda)
  lambda_hat[i] <- n/sum(sample)
}

var(lambda_hat)
lambda^2/n
```

## Example 1:

**Example:** The median $m$ of an Exponential($\lambda$) distribution satisfies P(X ≤ m) = 0.5. Solving $1 - e^{-\lambda m} = 0.5$ gives $m = ln(2) / \lambda$. This suggests an estimator for $\lambda$ based on the median: $\widehat{\lambda} = \frac{ln(2)}{m}$. Finding the analytical variance of $\widehat \lambda$ is complicated. Finding the sampling distribution of $m$ is complicated, and finding the non-linear transformation is also complicated.

Use simulation to see whether $\frac{ln(2)}{m}$ (a) is unbiased and (b) achieves the CRLB

```{r}
# your code here
```

# Example from Notes10: Cauchy vs Normal

## Running Means

The code below creates the "running mean" graph from the textbook for the Cauchy distribution. In your groups, talk through the steps and ask if you have questions. (It's a little different than other simulations that we've seen!) Run the code chunk a couple of times to get a sense of the behavior of the mean of a Cauchy distribution.

```{r}
n <- 10000
running_mean <- numeric(n)
sample <- rcauchy(n)

for(i in 1:n){
  running_mean[i] <- mean(sample[1:i])
}

ggplot() + 
  geom_line(aes(x = 1:n, y = running_mean)) 
```

Now, use the code chunk below to do the same "running mean" example for a Normal(0,1) distribution. What do you notice? How is it the same/different from the Cauchy example above? 

```{r}
# your code here
```

## Variance of sample means

The code chunk below runs a simulation comparing the variance of the sample mean of a Normal(0,1) distribution to the variance of the sample mean for a Cauchy distribution. Run it a few times to get a sense of the behavior, then try larger values of n. What do you notice? What does this mean about the *consistency* of the estimators?

```{r}
n <- 20
n_sims <- 1000
sample_mean_norm <- numeric(n)
sample_mean_cauchy <- numeric(n)

for(i in 1:n_sims){
  samp1 <- rnorm(n)
  samp2 <- rcauchy(n)
  sample_mean_norm[i] = mean(samp1)
  sample_mean_cauchy[i] = mean(samp2)
}

var(sample_mean_norm)
var(sample_mean_cauchy)
```
