---
title: "09: Efficiency and CRLB"
author: "Prof Amanda Luby"
subtitle: "Stat250 S25"
callout-appearance: minimal
knitr:
    opts_chunk: 
      dev: "ragg_png"
      echo: false
      warning: false
      message: false
format:
  pdf:
    include-in-header: 
       - "../preamble.tex"
    toc: false
    number-sections: true
    colorlinks: true
    geometry:
      - top=1in
      - left=1in
      - right=1in
      - bottom=1in
      - heightrounded
    mainfont: "Linux Libertine"
    sansfont: "Linux Libertine"
    mathfont: "Libertinus Math"
    monofont: "Monaco"
fontsize: 11pt
---


```{r, echo = FALSE}
library(tidyverse)
library(patchwork)
library(Sleuth3)
library(mosaic)
library(infer)
```

# Recap

- Observe data $X_1, ..., X_n \sim F_x(x|\theta)$, where $\theta$ is unknown

- Goal: *Estimate* $\theta$ based on the values of $X_i$ by formulating an *estimator* $\widehat \theta$

- One technique is to use the *maximum likelihood estimator*

- A second technique is to use the *method of moments estimator*

- There are lots of other ways to come up with estimators.

- We can compare estimators by comparing their *bias*, *variance*, and *mean squared error*

- Today: is there a way to know if we've found an "optimal" estimator?

# Comparing unbiased estimators 

::: callout-note
## Efficiency

For two unbiased estimators, $\widehat \theta_1$ is more **efficient** than $\widehat \theta_2$ if 

\vspace{.5in}
:::

**Example:** Why efficiency matters

We now have two unbiased estimators for $\theta$ in a Unif($0, \theta$) distribution. $\hat{\theta}_{MoM}=$ ____________ and $\hat{\theta_3}=$ __________________. 

(a) If $n=10$, what is the expectation and variance of each of these estimators?

\vspace{2.5in}

(b) What $n$ is needed for $\hat{\theta_{MoM}}$ to reach the variance of $\hat{\theta_3}$

\vspace{2.5in}

# Can we find a better estimator?

Is there another unbiased estimator with smaller variance?

::: callout-note
## Cramer-Rao Lower Bound

If $X_1, ..., X_n$ are an iid sample from a distribution with pdf $f(x|\theta)$, then any unbiased estimator $\hat\theta$ of $\theta$ satisfies: 

$$V(\hat{\theta}) \ge \frac{1}{n I(\theta)}$$

where $I(\theta)$ is the **Fisher Information** of $X_i$
:::

If the variance of an estimator is equal to the CRLB, then there is *no other unbiased estimator with more precision*

::: callout-note
## Fisher Information

The **Fisher Information** of an observation $X$ is
$$I(\theta) = E[(\frac{d}{d\theta} \ln f(x|\theta))^2] =$$

*provided the domain of $X$ does not depend on $\theta$ (and a few other regularity conditions)
:::

## Intuition: Fisher information

$$I(\theta) = E_X[(\frac{d}{d\theta} \ln f(x|\theta))^2] = -   E_X(l''(\theta))$$

- $l(\theta)$ is the log-likelihood function
- $l'(\theta)$ gives the slope of the log-likelihood at any point, $l''(\theta)$ gives the curvature at any given point
- The Fisher information "averages out" over $X$ and summarizes the overall curvature of the log-likelihood function

```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 2

p1 <- ggplot() + 
  geom_function(fun = \(x) log(dnorm(x))) + 
  xlim(c(-3,3)) + 
  labs(
    x = expression(theta),
    y = expression(ln(X,theta))
  )

p2 <- ggplot() + 
  geom_function(fun = \(x) log(dbeta(x, shape1 = 3, shape2=3))) + 
  xlim(c(0,1)) + 
  labs(
    x = expression(theta),
    y = expression(ln(X,theta))
  )


p1 + p2 & theme_bw()
```

## Practice with CRLB

**Example:**  Find the *Fisher information* for $Y$, where $X \sim Exp(\lambda)$

\vspace{3in}

**Example:** Let $Y_1, ..., Y_n$ be $n$ Exp($\lambda$) random variables Let $\widehat\lambda = \frac{n}{\sum Y_i}$. How does Var($\widehat\lambda$) compare with the CRLB?

\vspace{3in}

## CRLB with simulation

**Example:** The median $m$ of an Exponential($\lambda$) distribution satisfies P(X ≤ m) = 0.5. Solving $1 - e^(-\lambda m) = 0.5$ gives $m = ln(2) / \lambda$. This suggests an estimator for $\lambda$ based on the median: $\widehat{\lambda} = \frac{ln(2)}{m}$. Finding the analytical variance of $\widehat \lambda$ is complicated. Finding the sampling distribution of $m$ is complicated, and finding the non-linear transformation is also complicated.

Use simulation to see whether $\frac{ln(2)}{m}$ (a) is unbiased and (b) achieves the CRLB
