---
title: Intro to Bayesian Inference
subtitle: "Day 27"
title-slide-attributes:
  data-background-gradient: "linear-gradient(to right, #46337e, #440154)"
  data-slide-number: none
author: "Prof Amanda Luby"
format: 
  revealjs:
    incremental: true
    scrollable: true
editor: 
  markdown: 
    wrap: 72
---

```{r setup}
#| include: false

library(tidyverse)
library(countdown)

library(ggformula)
library(openintro)
library(patchwork)
library(fivethirtyeight)
library(dplyr)
library(tidyr)
library(fontawesome)
library(gt)

library(gridExtra)
library(tidyverse)
library(knitr)
library(mosaic)
library(infer)
library(kableExtra)
library(latex2exp)

library(plotly)
library(ggthemes)
library(janitor)
library(bayesrules)
# #440154
knitr::opts_chunk$set(echo = TRUE,
                  message = FALSE,
                  warning = FALSE)

slides_theme = theme_minimal(base_family = "serif", base_size = 24) +
  theme(plot.background = element_rect(fill = "#f0f1eb", colour = NA))
  

theme_set(slides_theme)

data("happy", package = "ggmosaic")
happy2018 <- filter(happy, year == 2018) %>%
  drop_na(happy, finrela)
yt <- 0
```

# A Bayesian "personality quiz"

:::{.footer}
Adapted from *Bayes Rules! An Introduction to Bayesian Modeling with R* 
:::



## Question 1

When flipping a fair coin, we say that "the probability of flipping Heads is 0.5." How do you interpret this probability?

::: nonincremental
1. If I flip this coin over and over, roughly 50% will be Heads.

2. Heads and Tails are equally plausible.

3. Both a and b make sense.
:::

## Question 2


An election is coming up and a pollster claims that "candidate A has a 0.9 probability of winning." How do you interpret this probability?

::: nonincremental
1. If we observe the election over and over, candidate A will win roughly 90% of the time.

2. Candidate A is much more likely to win than to lose.

3. The pollster's calculation is wrong. Candidate A will either win or lose, thus their probability of winning can only be 0 or 1.
:::

## Question 3


Consider two claims. 

::: nonincremental
- Lou claims that he can predict the outcome of a coin flip. To test his claim, you flip a fair coin 10 times and he correctly predicts all 10. 
- Kavya claims that she can distinguish natural and artificial sweeteners. To test her claim, you give her 10 sweetener samples and she correctly identifies each. 
:::

In light of these experiments, what do you conclude?

::: nonincremental
1. You're more confident in Kavya's claim than Lou's claim.

2. The evidence supporting Lou's claim is just as strong as the evidence supporting Kavya's claim.

:::

## Question 4


Suppose that during a recent doctor's visit, you tested positive for a very rare disease. If you only get to ask the doctor one question, which would it be?

::: nonincremental
1. What's the chance that I actually have the disease?

2. If in fact I don't have the disease, what's the chance that I would've gotten this positive test result?
:::

## Tally your points


::::{.columns }
:::{.column .nonincremental width="50%"}
Question 1:


- 1 = 1 points
- 2 = 3 points
- 3 = 2 points


Question 2:

- 1 = 1 points
- 2 = 3 points
- 3 = 1 points
:::
:::{.column .nonincremental width="50%"}
Question 3:

- 1 = 3 points
- 2 = 1 points
<br>
<br>

Question 4:


- 1 = 3 points
- 2 = 1 points
:::
::::



## What does your score mean?

- 4-5 $\rightarrow$ you're more of a frequentist thinker 

- 6-8 $\rightarrow$ you see the merit in both (a pragmatist?)

- 9-12 $\rightarrow$ you're more of a Bayesian thinker 




##
### Question 1: Interpreting probability


When flipping a fair coin, we say that "the probability of flipping Heads is 0.5." How do you interpret this probability?

::: nonincremental
1. (Frequentist) If I flip this coin over and over, roughly 50% will be Heads.

2. (Bayesian) Heads and Tails are equally plausible.

3. Both a and b make sense.
:::


##
### Question 2: Interpreting probability


An election is coming up and a pollster claims that "candidate A has a 0.9 probability of winning." How do you interpret this probability?

::: nonincremental
1. (Frequentist) If we observe the election over and over, candidate A will win roughly 90% of the time.

2. (Bayesian) Candidate A is much more likely to win than to lose.

3. (Rabid frequentist) The pollster's calculation is wrong. Candidate A will either win or lose, thus their probability of winning can only be 0 or 1.
:::


##
### Question 3: Balancing prior info and observed data


:::{style="font-size: 85%;"}
Consider two claims. 

::: nonincremental
- Lou claims that he can predict the outcome of a coin flip. To test his claim, you flip a fair coin 10 times and he correctly predicts all 10. 
- Kavya claims that she can distinguish natural and artificial sweeteners. To test her claim, you give her 10 sweetener samples and she correctly identifies each. 
:::

In light of these experiments, what do you conclude?

::: nonincremental
1. (Bayesian) You're more confident in Kavya's claim than Lou's claim.

2. (Frequentist) The evidence supporting Lou's claim is just as strong as the evidence supporting Kavya's claim.
:::

:::

##
### Question 4: Asking questions


Suppose that during a recent doctor's visit, you tested positive for a very rare disease. If you only get to ask the doctor one question, which would it be?

::: nonincremental

1. (Bayesian) What's the chance that I actually have the disease?

2. (Frequentist) If in fact I don't have the disease, what's the chance that I would've gotten this positive test result?
:::



# A first Bayesian example

## Example: 

A Des Moines register poll a few days before the 2024 Presidential election showed Kamala Harris with 51.61% of the 2-party vote share in a poll of $n=808$ likely voters. This poll result received a lot of buzz, because in the two months leading up to this poll, this proportion was estimated between 44.7 and 47.7. 

- Nov 2: 51.5% (x = 417; n=808)
- Nov 2: 44.7% (x = 358; n=800)
- Oct 2: 46.8% (x=281; n=600)
- Sept 15: 47.7% (x=382; n=800)

## Overview of the Bayesian method

1. Choose (or elicit) a probability distribution to express the pre-data belief about the parameter of interest, $\theta$.


2. Choose a model for the data given $\theta$.


3. Observe data, $Y_1, \ldots, Y_n$.


4. Update the belief about $\theta$ by combining the prior belief and the data.


5. Draw inferences using this updated belief about $\theta$.

##
### Overview of the Bayesian method - math



1. Assume a **prior distribution:** $\theta \sim f_\theta(\theta)$


2. Choose a **likelihood function** for the data: $Y|\theta \sim f_{y|\theta}(y|\theta)$


3. Observe data, $Y_1, \ldots, Y_n$.


4. Find the **posterior distribution** of $\theta$ given the data: $\theta \mid Y \sim f_{\theta \mid Y}(\theta \mid Y)$


5. Draw inferences with $f_{\theta|Y}$


## Choosing a data model

- **Likelihood**: a model for our data $X_1, \dotsc, X_n$

-  $X_i$ is drawn from a population/distribution with pdf/pmf $$X_i \sim f(x_i | \theta)$$

- The **joint** probability model for all $n$ data values (likelihood function)
$$f(x_1, \dotsc, x_n \mid \theta) = \prod_{i=1}^n f(x_i | \theta)$$

- Our inference goal will be to derive a model our unknown parameter(s) $\theta$ that is informed by our observed data values 

## Choosing a prior

- **Prior**: we give $\theta$ an initial probability model that reflects our beliefs about $\theta$ *prior* to observing our data
$$\theta \sim f(\theta)$$

- The model $f(\theta)$ can be either a pdf or pmf 

- and $f(\theta)$ should be defined on reasonable values of $\theta$
  - e.g. $\theta = p$ is a probability of success: then $f$ could be Unif[0,1]
  - e.g. $\theta = \sigma^2$ from a half-normal: then $f$ could be a Gamma model

## Finding the posterior


- **Posterior**: update our prior $f$ to reflect the information about $\theta$ that is provided by our data $x_1, \dotsc, x_n$
$$\theta \mid  x_1, \dotsc, x_n \sim f(\theta \mid x_1, \dotsc, x_n)$$

- We compute our conditional posterior distribution using Bayes theorem:
$$f(\theta \mid  x_1, \dotsc, x_n) = \dfrac{f(\theta, x_1, \dotsc, x_n)}{f(x_1, \dotsc, x_n)} = \dfrac{f(\theta)f(x_1, \dotsc, x_n \mid \theta)}{f(x_1, \dotsc, x_n)}$$

- the first expression is our definition of conditional probability

- the second expression is "Bayes theorem" which tells us (in words):
$$\textrm{posterior} = \dfrac{\textrm{prior} \times \textrm{data }}{\textrm{marginal  of data}}$$

## Finding the posterior

$$f(\theta \mid  x_1, \dotsc, x_n)  = \dfrac{f(\theta)f(x_1, \dotsc, x_n \mid \theta)}{f(x_1, \dotsc, x_n)} = \dfrac{\textrm{prior} \times \textrm{data }}{\textrm{marginal  of data}}$$

- Recall from probability: We find a marginal probability by integrating "out" nuisance variables from a joint distribution
$$\textrm{marginal pdf of data} = f(x_1, \dotsc, x_n) = \int_{-\infty}^{\infty} f(\theta)f(x_1, \dotsc, x_n \mid \theta) d\theta$$

## Binomial example: prior

- Goal is to estimate the parameter:
$$p = \textrm{proportion of Iowa likely voters who will vote for Harris}$$

- **Prior**: A natural prior for a proportion is a uniform:
$$p \sim Unif[0,1] \textrm{ so that } f(p) = 1 \textrm{ for } 0 \leq p \leq 1$$

. . . 

```{r, echo=FALSE, fig.height=3}
library(ggplot2)
ggplot(data.frame(p=c(0,1)), aes(p)) + 
  stat_function(fun = dunif)  +
  ggtitle("Our prior beliefs about p") + 
  labs(y="prior density") 
```


## Binomial example: data

- **Data** 808 respondents who responded to the 2-party presidential vote question. 

- Our random sample $X_1, \dotsc, X_{808}$ is Bernoulli: 
$$f(x_1, \dotsc, x_n \mid p) = \prod_{i=1}^{808}p^{x_i}(1-p)^{(1-x_i)} = p^{\sum_{i=1}^{808}x_i}(1-p)^{808 - \sum_{i=1}^{808}x_i}$$

- or, an **equivalent** data model just models the total number of "success" (approves) $X = \sum_{i=1}^{808}X_i$ which is a Binomial count:
$$f(x \mid p) = \binom{808}{x}p^x(1-p)^{808-x}$$

## Binomial example: posterior

- **Posterior** We need to derive the following pdf for $p$:
$$f(p \mid  x)  = \dfrac{f(\theta)f(x \mid p)}{f(x)} = \dfrac{1 \times \binom{n}{x}p^x(1-p)^{n-x}}{\int_{0}^1 1 \times \binom{n}{x}p^x(1-p)^{n-x} dp} \textrm{ for } 0 \leq p \leq 1$$

## Binomial example: posterior

- The toughest part of any posterior calculation is usually the marginal integration:
$$f(x) = \int_{0}^1\binom{n}{x}p^x(1-p)^{n-x} dp$$

- Any terms not involving $p$ can come outside the integrand:
$$f(x) = \binom{n}{x} \int_{0}^1p^x(1-p)^{n-x} dp$$

- ...which still looks bad, until you remember the Beta distribution which tells us that a Beta pdf integrates to 1 over the interval [0,1]:
$$1 =  \int_{0}^1\dfrac{\Gamma(a + b)}{\Gamma(a) \Gamma(b)}p^{a-1}(1-p)^{b-1} dp$$

## Binomial example: posterior

- ...which allows us to solve our integration problem without actually doing the integral
$$ \dfrac{\Gamma(a)\Gamma(b)}{\Gamma(a + b)} = \int_{0}^1p^{a-1}(1-p)^{b-1} dp$$

- Letting $a = x + 1$ and $b=n - x + 1$, we have
$$f(x) = \binom{n}{x} \int_{0}^1p^x(1-p)^{n-x} dp = \binom{n}{x}  \dfrac{\Gamma(x + 1)\Gamma(n - x + 1)}{\Gamma(n + 2)}$$

## Binomial example: posterior

- ...then for $0 \leq p \leq 1$
$$f(p \mid  x)  = \dfrac{\binom{n}{x}p^x(1-p)^{n-x}}{\binom{n}{x}  \dfrac{\Gamma(x + 1)\Gamma(n - x + 1)}{\Gamma(n + 2)}} = \dfrac{\Gamma(n + 2)}{\Gamma(x + 1)\Gamma(n - x + 1)} p^x(1-p)^{n-x}$$

- Which is the pdf for Beta distribution:
$$p \mid x \sim Beta(x + 1, n - x + 1)$$

- Iowa data: our posterior distribution for $n=808$ and $x=417$ is
$$p \mid x = 417 \sim Beta(418, 392)$$

## Binomial example: posterior {.smaller}

:::: {.columns}

::: {.column .nonincremental width="40%"}
- Inferences about $p$ use the Beta(391, 611) distribution. 
    - What values of $p$ are most *probable*? (interval)
    - What is the expected value of $p$? (point estimate)
    - How much uncertainty do we have in the distribution of $p$? (SE)
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 6
library(ggplot2)
ggplot(data.frame(p=c(0,1)), aes(p)) + 
  stat_function(fun = dunif, aes(color="prior", linetype="prior"), show.legend=TRUE, size = 2)  +
    stat_function(fun = dbeta, args = list(shape1=390+1, shape2 = 1000-390 + 1), aes(color="posterior Beta(391,611)", linetype="posterior Beta(391,611)"), show.legend=FALSE, size = 2)  +
  guides(color=guide_legend("model"), linetype=guide_legend("model")) + 
  labs(y="density") + 
  scale_color_brewer(type = "qual", palette = 6) +
  theme(legend.position = "none") + 
  scale_color_brewer(type = "qual", palette = 2)
```
:::
::::

## Binomial example: posterior

$$p \mid x = 417 \sim Beta(418, 392)$$

- **Point Estimate**: $\hat{p}_{bayes} = E(p \mid x=417) = \dfrac{418}{418 + 392} = \dfrac{418}{810} = 0.5160$

- **Variation**: $SD(p \mid x=417) = \sqrt{V(p \mid x=417)} = \sqrt{\dfrac{418(392)}{(418 + 392)^2(418 + 392 +1)}} = 0.0003$

- **Credible Interval for $p$**: A common choice of $100(1-\alpha)$% interval estimates is to use the $\alpha/2$ and $1-\alpha/2$ quantiles of the posterior

. . . 

```{r}
qbeta(c(.025,.975),418,392)  # 95% credible interval for p
```

## Binomial example: posterior

- What if we used classical frequentist methods instead of our Bayesian model?


- **Bayes Point Estimate of $p$**: Our posterior expectation for $p$ is
$\hat{p}_{bayes} =0.5160$


- **Frequentist Point Estimate of $p$**: MLE of $p$ is
$$\hat{p}_{MLE} = \dfrac{x}{n} = \dfrac{417}{808} = 0.5161$$

## Binomial example: posterior

- Why are the Bayes estimate and MLE so close?

- Rewrite $\hat{p}_{bayes}$:
$$\hat{p}_{bayes} = \dfrac{x+1}{n + 2} = \dfrac{n}{n+2}\dfrac{x}{n} + \dfrac{2}{n+2}\dfrac{1}{2} = \dfrac{n}{n+2}\hat{p}_{MLE} + \dfrac{2}{n+2}E(p)$$

- The Bayes estimate is a *weighted average* of the MLE and prior expectation

- The Unif[0,1] prior is like adding 2 units to the sample: 1 success and 1 failure so we have a prior expectation of 50% for $p$
    - weight $n/(n+2)$ favors the likelihood when $n > > 2$ (data size is much bigger than prior sample addition)
    - weight $2/(n+2)$ gives more weight to the prior mean when $n$ is closer to 2 


## Binomial example: posterior


- **Bayes SD**: $\sqrt{V(p \mid x=390)} =  0.01754$

- **Frequentist SE**: The sample proportion SE is
$$SE(\hat{p}) = \sqrt{\dfrac{\hat{p}(1-\hat{p})}{n}} = \sqrt{\dfrac{0.39(1-0.39)}{1000}}=0.01758$$
Same as the Bayes SD up to 4 decimal spots!

- We can write the posterior variance as 
$$V(p \mid x) = \dfrac{(x+1)(n-x+1)}{(n+2)^2(n+2 +1)} = \dfrac{\hat{p}_{bayes}(1-\hat{p}_{bayes})}{n+2 +1}$$
which is similar in form as the frequentist SE. 

## Binomial example: posterior

**Bayes Credible Interval for $p$**: There is a 95% **probability** that the Harris two-party vote share is between 48.16% and 55.04%. 
```{r}
qbeta(c(.025,.975),418,392)  # 95% credible interval for p
```

. . . 

**Frequentist Confidence Interval for $p$**: I am 95% **confident** that he Harris two-party vote share is between 48.1% and 55.1%. 

```{r}
prop.test(417,808)$conf  # 95% confidence interval for p
```

. . . 

Intervals contain very similar values, but are interpreted very differently!


## What if we chose a different prior? 

In the three polls leading up to the Des Moines register poll, Harris' 2-party vote share hovered between 44.7 and 47.7 percent. What if my prior distribution took that information into account? 

```{r}
#| echo: false
plot_beta(1,1) + plot_beta(140, 160)
```

## What if we chose a different prior?

- A more flexible prior for $p$ is a Beta( $\alpha, \beta$ ) distribution:
$$f(p) = \dfrac{\Gamma(\alpha + \beta)}{\Gamma(\alpha)\Gamma(\beta)} p^{\alpha - 1}(1-p)^{\beta-1} \textrm{ for } 0 \leq p \leq 1$$

- The posterior then looks like:
$$f(p \mid  x)   = \dfrac{\dfrac{\Gamma(\alpha + \beta)}{\Gamma(\alpha)\Gamma(\beta)} p^{\alpha - 1}(1-p)^{\beta-1} \times \binom{n}{x}p^x(1-p)^{n-x}}{f(x)} \\
\ \ \ \ \ \ \ \ \ \ \ \  = \dfrac{\dfrac{\Gamma(\alpha + \beta)}{\Gamma(\alpha)\Gamma(\beta)} \binom{n}{x} }{f(x)}p^{x + \alpha - 1}(1-p)^{n - x + \beta -1}  \textrm{ for } 0 \leq p \leq 1$$

## Beta-Binomial model

- The **kernel** of the posterior (the part that involves $p$) looks like:
$$f(p \mid  x) \propto  p^{x + \alpha - 1}(1-p)^{n - x + \beta -1}  \textrm{ for } 0 \leq p \leq 1$$

- Since $f(p \mid x)$ must integrate to 1 over [0,1], the kernel uniquely identifies the pdf as a Beta( $x + \alpha, n-x+\beta$ )
    - which means the **normalizing constant** for this kernel is
$$\dfrac{\Gamma(n + \alpha + \beta)}{\Gamma(x + \alpha)\Gamma(n-x+\beta)} = \dfrac{\dfrac{\Gamma(\alpha + \beta)}{\Gamma(\alpha)\Gamma(\beta)} \binom{n}{x} }{f(x)}$$
    -  we just need to look at
$$f(\theta \mid x_1, \dotsc, x_n) \propto f(\theta)f(x_1, \dotsc, x_n \mid \theta)$$
    
    
## Beta-Binomial model

- Data: $X \mid p \sim Binom(n,p)$
- Prior: $p \sim Beta(\alpha, \beta)$ 
- Prior Expectation: $E(p) = \dfrac{\alpha}{\alpha + \beta}$


- Posterior: $p \mid x \sim Beta(x + \alpha, n-x + \beta)$
- Posterior Expectation: $E(p \mid x) = \dfrac{x + \alpha}{n + \alpha + \beta} = \dfrac{n}{n+\alpha + \beta}\hat{p}_{MLE} + \dfrac{\alpha + \beta}{n+\alpha + \beta}E(p)$


- The prior is like adding $\alpha$ succeses and $\beta$ failures to the data. 
    - Larger values mean more weight given to the prior (success and/or failures)

## Beta-Binomial model: flat prior

```{r,  echo=FALSE}
library(ggplot2)
plot_beta_binomial(1,1,y=417, n=808)

```

## Beta-Binomial model: somewhat informative prior

```{r,  echo=FALSE}
library(ggplot2)
plot_beta_binomial(140,160,y=417, n=808)
```

## Beta-Binomial model: extremely informative prior

```{r,  echo=FALSE}
library(ggplot2)
plot_beta_binomial(1021,1179,y=417, n=808)
```

## Beta-Binomial model

For $n = 808$ and $x=417$ successes:
```{r, echo=FALSE}
summarize_beta_binomial(1, 1, 417, 808)
summarize_beta_binomial(140, 160, 417, 808)
summarize_beta_binomial(1021, 1179, 417, 808)
```


As our prior contains a bigger "sample size", we see 
  - a posterior mean pulled more towards the prior mean
  - a reduction in the posterior standard deviation

## Comparing intervals

If we compare credible intervals for the 3 posteriors: 

```{r}
qbeta(c(.025,.975),418,392)  
qbeta(c(.025,.975),557,551)  
qbeta(c(.025,.975),1438,1570)  
```

*Note:* on election day, Harris received 43.27% of the two-party vote share
