---
title: "26: Inference for SLR"
author: "Prof Amanda Luby"
subtitle: "Stat250 S25"
callout-appearance: minimal
knitr:
    opts_chunk: 
      dev: "ragg_png"
      echo: false
      warning: false
      message: false
format:
  pdf:
    include-in-header: 
       - "../preamble.tex"
    toc: false
    number-sections: true
    colorlinks: true
    geometry:
      - top=1in
      - left=1in
      - right=1in
      - bottom=1in
      - heightrounded
    mainfont: "Linux Libertine"
    sansfont: "Linux Libertine"
    mathfont: "Libertinus Math"
    monofont: "Monaco"
fontsize: 11pt
---

```{r}
library(tidyverse)
library(ggformula)
library(ggthemes)
library(knitr)
library(mosaic)
library(patchwork)
library(palmerpenguins)
library(ggridges)
library(janitor)
library(gt)
theme_set(theme_minimal())
data("happy", package = "ggmosaic")
happy2018 <- filter(happy, year == 2018) %>%
  drop_na(happy, finrela)
yt <- 0
```

```{r}
bullets <- readxl::read_excel("../17-two-sample-ci/monson.xlsx", sheet = 2) |>
  group_by(ExamID) |>
  summarize(
    er_bb = mean(`Score` %in% c("Incorrect-actually Match(gt)", "Incorrect-actually Non Match(gt)")),
    inc_bul = mean(Score %in% c("Match(gt) marked Inconclusive", "Non-Matched(gt) marked Inconclusive"))
  )

cc <- readxl::read_excel("../17-two-sample-ci/monson.xlsx", sheet = 3) |>
  group_by(ExamID) |>
  summarize(
    er_cc = mean(`Score` %in% c("Incorrect-actually Match(gt)", "Incorrect-actually Non Match(gt)")),
    inc_cart = mean(Score %in% c("Match(gt) marked Inconclusive", "Non-Matched(gt) marked Inconclusive"))
  )

set.seed(0508)
firearms <- bullets |>
  left_join(cc, by = "ExamID") 
```


# SLR Model Recap 

$$Y_i | x_i \sim N(\beta_0 + \beta_1 x_i, \sigma^2)$$

Key features:

\vspace{1.75in}

All three of these data sets were generated from the SLR model:

```{r, echo=FALSE}
n<- 3000
set.seed(7712)
e <- rnorm(n,0,2)
x1 <- runif(n,0,20)
y1 <- 2 + 1*x1 + e
x2 <- rexp(n,rate=.3)
y2 <- 2 + 1*x2 + e
x3 <- rnorm(n,10,3)
y3 <- 2 + 1*x3 + e
df <- data.frame(model = rep(1:3,each=n),x=c(x1,x2,x3),y=c(y1,y2,y3))
```

```{r, echo=FALSE}
#| fig-width: 6
#| fig-height: 2
ggplot(df, aes(x=x,y=y)) + 
  geom_point(size = .25) + 
  geom_smooth(method="lm", se=FALSE) + 
  facet_wrap(~model, scales="free")
```

But only data 3 has (marginally) normally distributed responses:

```{r, fig.height=5, fig.width=9, echo=FALSE}
#| fig-width: 6
#| fig-height: 2
ggplot(df, aes(sample=y)) + 
  geom_qq(size = .5) + stat_qqline(color="darkgray")+
  facet_wrap(~model, scales="free") + 
  ggtitle("Normal QQ plots of all Y's")
```

Let's look at the Y values for all cases with $x \approx 7$:

```{r, fig.height=5, fig.width=9, echo=FALSE}
#| fig-width: 6
#| fig-height: 2
df2 <- df %>% filter(x > 6.5, x < 7.5)
ggplot(df, aes(x=x,y=y)) + 
  geom_point(size = .25) + geom_point(size = .25, data=df2, aes(x,y), color="darkorange") + 
  geom_smooth(method="lm", se=FALSE) + 
  facet_wrap(~model, scales="free")
```

The Y values are (conditionally) normally distributed for with $x \approx 7$ 

```{r, fig.height=5, fig.width=9, echo=FALSE}
#| fig-width: 6
#| fig-height: 2
ggplot(df2, aes(sample=y)) + 
  geom_qq(size = .5) + stat_qqline(color="darkgray")+
  facet_wrap(~model, scales="free") + 
  ggtitle("Normal QQ plots of all Y's")
```

# Inference for coefficients

::: callout-note
## Maximum Likelihood Estimators for SLR

$$\hat \beta_0 = \bar{Y} - \hat\beta_1 \bar{x}$$

$$\hat \beta_1  = \frac{\sum(x_i - \bar{x})(Y_i - \bar{Y})}{\sum (x_i - \bar{x})^2}$$

$$ \hat\sigma^2 = \frac{\sum e_i^2}{n}$$
:::

Properties of the MLE's for SLR: 

1. $\hat \beta_0$ and $\hat \beta_1$ are Normally-distributed random variables

2. $\hat \beta_0$ and $\hat \beta_1$ are unbiased estimators

3. $V(\hat \beta_1) = \frac{\sigma^2}{\sum (x_i - \bar{x})^2}$

4. $V(\hat \beta_0) = \sigma^2 [\frac{1}{n} + \frac{\bar{x}^2}{\sum(x_i - \bar{x})^2}]$

5. $\hat \beta_1 , \bar{Y}$ and $\hat \sigma^2$ are mutually independent

6. $\frac{n \hat \sigma^2}{\sigma^2} \sim \chi^2_{n-2}$

7. $s^2 = \frac{n}{n-2} \hat \sigma^2$ is an unbiased estimator for $\sigma^2$


::: callout-note
## Test statistic for $\beta_1$

Let $(x_1, Y_1), (x_2, Y_2), ..., (x_n, Y_n)$ be a set of points satisfying $E(Y|X=x) = \beta_0 + \beta_1 x$ and let $S^2 = \frac{1}{n-2} \sum (Y_i - (\hat\beta_0 + \hat\beta_1 x_i))^2$. Then, 

$T = \frac{\hat\beta_1 - \beta_1}{S/\sqrt{\sum (x_i - \bar{x})^2}}$
:::

\vspace{2.5in}


**Example:**  We are interested in the average *inconclusive rate* (how often the firearms examiner cannot come to a definitive conclusion) for bullets compared to cartridge cases. Some results from a study are included below:

```{r}
#| fig-width: 4
#| fig-height: 2

ggplot(firearms, aes(x = inc_bul, y = inc_cart)) + 
  geom_point(size = .5) + 
  geom_smooth(method = "lm", se = FALSE)
```

```{r}
#| echo: true 
cb_mod <- lm(inc_cart ~ inc_bul, data = firearms)
summary(cb_mod)
```

What is a 95% confidence interval for $\beta_1$?

\vspace{1.5in}

# Inference for $\mu_{Y|x}$

**Example:** How do we find a confidence interval for the line? 

\vspace{4in}







# Inference for new data points

**Example:** If we observe a new examiner who was inconclusive 75% of time time on bullets, what would we predict for their inconclusive rate on cartridge cases? 


\vspace{4in}

```{r}
#| echo: true
#| collapse: true
#| comment: ">"
predict(cb_mod, newdata = data.frame(inc_bul = .75))
predict(cb_mod, newdata = data.frame(inc_bul = .75), interval = "confidence")
predict(cb_mod, newdata = data.frame(inc_bul = .75), interval = "prediction")
```

```{r}
#| fig-width: 4
#| fig-height: 2

ggplot(firearms, aes(x = inc_bul, y = inc_cart)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) + 
  geom_ribbon(data = broom::augment(cb_mod, interval = "prediction"), aes(ymin = .lower, ymax = .upper), fill = "darkorange",  alpha = .2) +
  geom_ribbon(data = broom::augment(cb_mod, interval = "confidence"), aes(ymin = .lower, ymax = .upper), fill = "steelblue",  alpha = .4) 
```
