---
title: "21: Likelihood Ratio Tests"
author: "Prof Amanda Luby"
subtitle: "Stat250 S25"
callout-appearance: minimal
knitr:
    opts_chunk: 
      dev: "ragg_png"
      echo: false
      warning: false
      message: false
format:
  pdf:
    include-in-header: 
       - "../preamble.tex"
    toc: false
    number-sections: true
    colorlinks: true
    geometry:
      - top=1in
      - left=1in
      - right=1in
      - bottom=1in
      - heightrounded
    mainfont: "Linux Libertine"
    sansfont: "Linux Libertine"
    mathfont: "Libertinus Math"
    monofont: "Monaco"
fontsize: 11pt
---

```{r}
library(tidyverse)
library(ggformula)
library(ggthemes)
library(knitr)
library(mosaic)
library(patchwork)
library(palmerpenguins)
library(ggridges)
theme_set(theme_minimal())
```

Most of the inference that we've covered so far 

**Example:** Suppose we observe $X_1, ..., X_9 \sim Exp(\theta)$ and are interested in testing $H_0: \theta = 9$ against $H_A: \theta = 10$. 

\vspace{1in}

```{r}
#| fig-width: 4
#| fig-height: 2
results <- tibble(
  h0 = numeric(1e4), 
  ha = numeric(1e4)
)
for(i in 1:1e4){
results$h0[i] <- sum(rexp(9, rate = 8))
results$ha[i] <- sum(rexp(9, rate = 10))
}

results |>
  pivot_longer(everything()) |> 
  ggplot(aes(x = value, fill = name)) + 
  geom_histogram(alpha = .7, col = "white", position = "identity") + 
  scale_fill_viridis_d(end = .8, option = "plasma") + 
  labs(
    x = expression(sum(X[i])),
    fill = "",
    y = ""
  )
```

::: callout-note
## Likelihood Ratio Test Statistic

\vspace{1in}
:::

\vspace{3in}

::: callout-note
## Neyman-Pearson Lemma

\vspace{2in}
:::

## LRT for Composite Hypotheses

::: callout-note
## Likelihood Ratio Test Statistic

\vspace{1in}
:::

**Example:** Suppose we observe $X_1, ..., X_9 \sim Exp(\theta)$ and are interested in testing $H_0: \theta \le 8$ against $H_A: \theta > 8$. 

\pagebreak


::: callout-note
## Wilk's Theorem

\vspace{1.5in}
:::

```{r}
#| fig-width: 4
#| fig-height: 2


glrt <- numeric(1e4)

for(i in 1:1e4){
  x <- rexp(9, 11)
  glrt[i] <- -2*log((8^9 * exp(-8*sum(x))) / ((9/sum(x))^9 *exp(-9)))
}
ggplot(tibble(glrt = glrt), aes(x = glrt)) +
  geom_histogram(col = "white", aes(y = ..density..)) + 
  stat_function(fun = dchisq, args = list(df = 2), color = "darkred", linewidth = 1) +
  labs(
    title = "Simulated GLRT T's compared to X(1)"
  )

```
