---
title: "25: Least Squares"
author: "Prof Amanda Luby"
subtitle: "Stat250 S25"
callout-appearance: minimal
knitr:
    opts_chunk: 
      dev: "ragg_png"
      echo: false
      warning: false
      message: false
format:
  pdf:
    include-in-header: 
       - "../preamble.tex"
    toc: false
    number-sections: true
    colorlinks: true
    geometry:
      - top=1in
      - left=1in
      - right=1in
      - bottom=1in
      - heightrounded
    mainfont: "Linux Libertine"
    sansfont: "Linux Libertine"
    mathfont: "Libertinus Math"
    monofont: "Monaco"
fontsize: 11pt
---

```{r}
library(tidyverse)
library(ggformula)
library(ggthemes)
library(knitr)
library(mosaic)
library(patchwork)
library(palmerpenguins)
library(ggridges)
library(janitor)
library(gt)
theme_set(theme_minimal())
data("happy", package = "ggmosaic")
happy2018 <- filter(happy, year == 2018) %>%
  drop_na(happy, finrela)
yt <- 0
```

**Example:**  Let's return to the Adelie penguins that we've seen before. The Long Term Ecological Research Network (LTER) has collected data on a group of Adelie penguins, including their body mass, flipper length, bill length, and bill depth. It’s relatively easy to measure their body mass, but harder to get accurate measurements of their flipper length. We generally expect heavier penguins will have longer flippers, but the researcher would like to know (a) how strong this relationship is and (b) the magnitude of this relationship.

```{r}
#| fig-width: 4
#| fig-height: 2

adelie <- penguins |> filter(species == "Adelie") |> drop_na(body_mass_g, flipper_length_mm)
ggplot(adelie, aes(x = body_mass_g, y = flipper_length_mm)) + 
  geom_point()

```

**Notation:** 

\vspace{1in}



::: callout-note
## Covariance

$$Cov[X, Y] = E[(X - \mu_X)(Y - \mu_Y)] = E[XY] - \mu_X \mu_Y$$
:::

::: callout-note
## Correlation

$$\rho(X,Y) =$$

:::

```{r}
#| fig-width: 4
#| fig-height: 2

z_score <- function(x){
  (x - mean(x)) / sd(x)
}
ggplot(adelie, aes(x = z_score(body_mass_g), y = z_score(flipper_length_mm))) + 
  geom_point() + 
  labs(x = "Body Mass Z Score",
       y = "Flipper Z Score") + 
  geom_hline(yintercept = 0 , col = "darkorange") +
  geom_vline(xintercept = 0 , col = "darkorange") 
```

::: callout-warning
## Watch out! 

Correlation only measures the strength of **linear** functions

:::

```{r}
#| collapse: true
#| comment: "#>"
#| fig-width: 3
#| fig-height: 2
#| echo: true

x <- -5:5
y <- x^2
cor(x, y)
gf_point(y ~ x)
```

::: callout-warning
## Watch out! 

Correlation is **sensitive to outliers**

:::

```{r}
#| collapse: true
#| comment: "#>"
#| fig-width: 6
#| fig-height: 2

p1 <- qplot(x3, y3, data = anscombe) +
  xlab("x") + ylab("y") + scale_y_continuous(limits=c(5,13)) +
  annotate("text", label = "r = 0.816", x = 6, y = 12, size = 3) +
  theme_light()

p2 <- qplot(x3, y3, data = subset(anscombe, y3 < 9)) +
  xlab("x") + ylab("y") + scale_y_continuous(limits=c(5,13)) +
  annotate("text", label = "r = 0.999", x = 6, y = 12, size = 3) +
  theme_light()

p1 + p2
```

# Least Squares Line

- Goal: predict flipper length based on body mass using a linear function: $$\hat y = \hat \alpha + \hat b x_i$$
- Approach: minimize the **residual sum of squares**: 

\vspace{.5in}

In math: 

\vspace{1.5in}

In R: 

```{r}
#| collapse: true
#| comment: "#>"
#| echo: true


adelie_fit <- lm(flipper_length_mm ~ body_mass_g, data = adelie)
```

To print the least squares estimates:

```{r}
#| collapse: true
#| comment: "#>"
#| echo: true

adelie_fit
```

**Interpretations:**

$\hat a$:

\vspace{1in}

$\hat b$: 

# "Nonlinear" Least Squares

Obviously, not every relationship can be adequately described by a straight line. BUT linear models are very "nice" with "easy" solutions (as we saw above). Luckily, we can "linearize" many nonlinear relationship by transforming the X or Y variable.

**Example:** We suspect that the true relationship between variables $X$ and $Y$ is described by $y = a e^{bx}$. What are the transformations of $X$ and $Y$ that "linearize" this relationship? 

\vspace{1in}


**Exercise:** Fill in the following table to show that all of these nonlinear relationships can be expressed as linear functions of transformations of the original variables.

| True Relationship              | Transformation of Y | Transformation of X |
|--------------------------------|---------------------|---------------------|
| $y = a + b x^2$                |                     |                     |
|                                |                     |                     |
| $y = a e^{bx}$                 |                     |                     |
|                                |                     |                     |
| $y = ax^b$                     |                     |                     |
|                                |                     |                     |
| $y = \frac{1}{1 + \exp(a+bx)}$ |                     |                     |
|                                |                     |                     |
| $y= \frac{1}{a+bx}$            |                     |                     |
|                                |                     |                     |
| $y = \frac{x}{a+bx}$           |                     |                     |
|                                |                     |                     |
| $y = 1-e^{-x^b/a}$             |                     |                     |

# Simple Linear Regression Model

Everything we’ve talked about up until this point has not used any statistical properties at all: there have been no probability distributions, expectations, independence assumptions, etc. We’ve gone about “fitting curves” as a purely geometric exercise.

**Example:** Suppose we observe an Adelie penguin that weighs 4,000g. 

- What is the prediction for that penguin's flipper length? 
- Would every 4,000g penguin have that flipper length? 

::: callout-note
## Simple Linear Regression Model

* Least Squares only assumes that there is a linear relationship between $x$ and $y$

* Additional assumptions are needed to understand the uncertainty of our predictions

* The SLR model can be written in a few forms

    + $Y_i = \alpha + \beta x_i + \varepsilon_i$ where $\varepsilon\overset{iid}{\sim} N(0, \sigma^2)$

    + $Y_i \overset{iid}{\sim} N(\alpha + \beta x_i, \sigma^2)$
:::

```{r}
#| collapse: true
#| comment: "#>"
#| fig-width: 4
#| fig-height: 3


set.seed(0)
dat <- data.frame(x=(x=runif(10000, 0, 50)),
                  y=rnorm(10000, 10*x, 100))

## breaks: where you want to compute densities
breaks <- seq(0, max(dat$x), len=5)
dat$section <- cut(dat$x, breaks)

## Get the residuals
dat$res <- residuals(lm(y ~ x, data=dat))

## Compute densities for each section, flip the axes, add means of sections
## Note: densities need to be scaled in relation to section size (2000 here)
dens <- do.call(rbind, lapply(split(dat, dat$section), function(x) {
  d <- density(x$res, n=5000)
  res <- data.frame(x=max(x$x)- d$y*1000, y=d$x+mean(x$y))
  res <- res[order(res$y), ]
  ## Get some data for normal lines as well
  xs <- seq(min(x$res), max(x$res), len=5000)
  res <- rbind(res, data.frame(y=xs + mean(x$y),
                               x=max(x$x) - 1000*dnorm(xs, 0, sd(x$res))))
  res$type <- rep(c("empirical", "normal"), each=5000)
  res
}))
dens$section <- rep(levels(dat$section), each=10000)

ggplot(dat, aes(x, y)) +
  geom_point(size = 0.1, alpha = 0.3) +
  geom_smooth(method="lm", fill=NA, lwd=2) +
  geom_path(data=dens[dens$type=="normal",], aes(x, y, group=section),
            color="orange", lwd=1.1) +
  geom_vline(xintercept=breaks, lty=2) +
  ggtitle("SLR model")
```

**LINE** Mnemonic: 

1. **Linearity**: $E(Y_i|X = x_i) = \mu_i = \alpha + \beta x_i$

1. **Independence**: $\varepsilon_1, \ldots, \varepsilon_n$ are independent

1. **Normal error terms**: $\varepsilon_i \sim N(0, \sigma^2)$

1. **Equal error variance**: $Var(\varepsilon_1) = \cdots = Var(\varepsilon_n) = \sigma^2$

*Note:* This model also assumes that the $x$ variables are **fixed**, which is why they get little x's instead of big X's

