---
title: "16: Comparing confidence interval procedures"
---

The {nycflights23} R package contains a dataset called `flights`. This dataset contains all flights in and out of NYC-area airports in 2023. Since the dataset has **all** flights, we can treat it as a population. We're interested in the average **departure delay** of **flights departing from NYC area airports**. 

```{r}
library(tidyverse)
library(nycflights23)

flights <- flights |>
  filter(origin %in% c("JFK", "EWR", "LGA")) |> # filter to departing flights from NYC
  drop_na(dep_delay) # drop observations with missing dep_delay variables

glimpse(flights)
```

::: callout-note
## Q1: EDA 

Make a histogram of `dep_delay`. This represents our population. What do you notice about the distribution? What is the true $\mu$ in this case?
:::


Let's explore the performance of our different confidence interval procedures. To do so, you'll first draw a *sample* from the `flights` dataset. You'll treat this as your data sample throughout the rest of the activity, and compare your confidence intervals to the true value that you found in Q1. 

::: callout-note
## Q2: Draw your sample 

Use the code below to draw a sample of size 40 from the `flights` dataset. Make sure to set a seed that you think is different from everyone else in the class! This is the "sample data" that you'll work with for the rest of this activity.
:::

```{r}
set.seed(1) # Replace "1" with a random seed
sample_index <- sample(1:nrow(flights), size = 40)
flights_sample <- flights[sample_index,]
dep_delay <- flights_sample$dep_delay
```

::: callout-note
## Q3: Construct your confidence intervals

Using your sample `dep_delay` from above, compute three confidence intervals: 

1. percentile bootstrap
2. CLT-based t
3. bootstrap t-based 

template code can be found in the handout for today
:::

when you're done, enter your lower and upper bounds [in this google sheet](https://docs.google.com/spreadsheets/d/1K3v67icpu-Q9vpYMcbCUsGJDJST2YPFGKtvgtmyIaQI/edit?usp=sharing)


::: callout-note
## Q4: (stretch goal, if time) Run coverage simulation

Try to set up a "simulation of simulations". Repeat your confidence interval procedure 100 times, on 100 different samples from `flights`, storing the upper and lower bounds for each confidence interval procedure as you go. 
:::
