---
title: "16: Comparing confidence intervals"
---

```{r}
library(tidyverse)
library(nycflights23)
glimpse(flights)
```

::: callout-note
## Q1: EDA 

Make a histogram of `dep_delay`. This represents our population. What do you notice about the distribution? What is the true $\mu$ in this case?
:::

::: callout-note
## Q2: Draw your sample 

Use the code below to draw a sample of size 40 from the `flights` dataset. Make sure to set a seed! This is the "sample data" that you'll work with for the rest of this activity.
:::

```{r}
set.seed(1) # Replace "1" with a random seed
sample_index <- sample(1:nrow(flights), size = 40)
flights_sample <- flights[sample_index,]
dep_delay <- flights_sample$dep_delay
```

::: callout-note
## Q3: Construct your confidence intervals

Using your sample `dep_delay` from above, compute three confidence intervals: 

1. percentile bootstrap
2. CLT-based t
3. bootstrap t-based 

template code can be found in the handout for today
:::

when you're done, enter your lower and upper bounds here: https://docs.google.com/spreadsheets/d/1K3v67icpu-Q9vpYMcbCUsGJDJST2YPFGKtvgtmyIaQI/edit?usp=sharing 

