---
title: "Efficiency & CRLB"
subtitle: "Day 09"
title-slide-attributes:
  data-background-gradient: "linear-gradient(to right, #46337e, #440154)"
  data-slide-number: none
author: "Prof Amanda Luby"
format: 
  revealjs:
    incremental: true
editor: 
  markdown: 
    wrap: 72
---

```{r setup}
#| include: false

library(tidyverse)
library(countdown)

library(ggformula)
library(openintro)
library(patchwork)
library(fivethirtyeight)
library(dplyr)
library(tidyr)
library(fontawesome)

library(gridExtra)
library(tidyverse)
library(knitr)
library(mosaic)
library(infer)
library(kableExtra)
library(latex2exp)

library(plotly)

# #440154
knitr::opts_chunk$set(echo = TRUE,
                  message = FALSE,
                  warning = FALSE)

slides_theme = theme_minimal(base_family = "serif", base_size = 24) +
  theme(plot.background = element_rect(fill = "#f0f1eb", colour = NA))
  

theme_set(slides_theme)
```

## Recap

- Observe data $X_1, ..., X_n \sim F_x(x|\theta)$, where $\theta$ is unknown
- Goal: *Estimate* $\theta$ based on the values of $X_i$ by formulating an *estimator* $\widehat \theta$
- One technique is to use the *maximum likelihood estimator*
- A second technique is to use the *method of moments estimator*
- We can compare estimators by comparing their *bias*, *variance*, and *mean squared error*.
- Today: is there a way to know if we've found an "optimal" estimator?

## Last time: Unif(0, $\theta$) distribution^[So $0 \le x \le \theta$, $f_x = \frac{1}{\theta}$, $F_x = \frac{x}{\theta}$, $E(X) = \frac{\theta}{2}$, and $V(X) = \frac{\theta^2}{12}$]

Your task is to compare the estimators 

$$\widehat{\theta}_{MLE} = X_\max \hspace{1in} \widehat{\theta}_{MoM} = 2\bar{X}$$ 

::: nonincremental
(a) What is the bias of each estimator?^[A helpful fact is that $f_{X_{max}}(x) = n[F(x)]^{n-1}f_X(x)$]
(b) What is the SE of each estimator?
(c) What is the MSE of each estimator?
(d) When does $\widehat{\theta}_{MLE}$ "beat"  $\widehat{\theta}_{MoM}$ in terms of MSE?
:::


## 

**(d) When does $\widehat{\theta}_{MLE}$ "beat"  $\widehat{\theta}_{MoM}$ in terms of MSE?**

MSE = "MSE factor" $\times \theta^2$

```{r}
#| echo: false


df <- data.frame(n = 1:10) %>%
  mutate(mle = 2/((n+2)*(n+1)), 
         mom = 1/(3*n)) %>%
  pivot_longer(2:3,names_to = "estimator", values_to = "mse factor" )

ggplot(df, aes(n,y = `mse factor`, color = estimator)) + 
  geom_point(aes(shape = estimator), size = 3) + geom_line(aes(linetype = estimator)) +
  labs(col = "",
       shape = "",
       linetype = "") + 
  scale_color_viridis_d(end = .8, option = "plasma") + 
  scale_x_continuous(breaks = 1:10)
```

## Example: Uniform($0,\theta$)

From last time: $E(\widehat \theta_{MLE}) = \frac{n}{n+1} \theta$ and $V(\widehat \theta_{MLE}) = [\frac{n}{(n+2)(n+1)^2}]$

::: nonincremental
(e) Since $\hat \theta_{MLE}$ beats $\hat \theta_{MoM}$ in terms of MSE but is biased, can we "unbias" the MLE? Call this third estimator $\hat\theta_3$

(f) Does $\hat \theta_3$ ever "beat" $\hat \theta_{MLE}$ in terms of MSE?
:::

## 

**(f) Does $\hat \theta_3$ ever "beat" $\hat \theta_{MLE}$ in terms of MSE?**

MSE = "MSE factor" $\times \theta^2$


```{r, echo = FALSE}
df <- data.frame(n = 1:10) %>%
  mutate(mle = 2/((n+2)*(n+1)), 
         mom = 1/(3*n), 
         mle_unbiased = 1/(n*(n+2))) %>%
  pivot_longer(2:4,names_to = "estimator", values_to = "mse factor" )
ggplot(df, aes(n,y = `mse factor`, color = estimator)) + 
  geom_point(aes(shape = estimator), size = 3) + geom_line(aes(linetype = estimator)) +
  labs(col = "",
       shape = "",
       linetype = "") + 
  scale_color_viridis_d(end = .8, option = "plasma") + 
  scale_x_continuous(breaks = 1:10)
```

## Moral of the story

We can "fix" bias, but it's harder to fix variance

## Comparing unbiased estimators: efficiency

::: callout-note
## Efficiency

For two unbiased estimators, $\widehat \theta_1$ is more **efficient** than $\widehat \theta_2$ if 

$$\text{Var}(\widehat \theta_1) < \text{Var}(\widehat \theta_2) \iff \text{SE}(\widehat \theta_1) < \text{SE}(\widehat \theta_2)$$
:::

## Example: why efficiency matters

We now have two unbiased estimators: 
$$\widehat \theta_{MoM}$\widehat \theta_{3} = \frac{n+1}{n} X_\max$$

(a) If $n=10$ what is the expectation and variance of each of these estimators?

## Example: why efficiency matters

(b) What $n$ would we need for $\widehat\theta_{MOM}$ to reach the variance of $\widehat\theta_3$?

# Can we find a "better" estimator?

. . . 

Is there another unbiased estimator with smaller variance?


## Comparing unbiased estimators: Cramer-Rao Lower Bound (CRLB)

. . . 

::: callout-note
## CRLB

If $X_1, ..., X_n$ are an iid sample from a distribution with pdf $f(x|\theta)$, then any unbiased estimator $\hat\theta$ of $\theta$ satisfies: 

$$V(\hat{\theta}) \ge \frac{1}{n I(\theta)}$$

where $I(\theta)$ is the **Fisher Information** of $X_i$
:::

. . . 

If the variance of an estimator is equal to the CRLB, then there is *no other unbiased estimator with more precision*

## Fisher Information

::: callout-note
## Fisher Information

The **Fisher Information** of a random variable $X$ is
$$I(\theta) = E[(\frac{d}{d\theta} \ln f(x|\theta))^2] = -   E(l''(\theta))$$

*provided the domain of $X$ does not depend on $\theta$ (and a few other regularity conditions)
:::

## Intuition: Fisher information

$$I(\theta) = E_X[(\frac{d}{d\theta} \ln f(x|\theta))^2] = -   E_X(l''(\theta))$$

- $l(\theta)$ is the log-likelihood function
- $l'(\theta)$ gives the slope of the log-likelihood at any point, $l''(\theta)$ gives the curvature at any given point
- The Fisher information "averages out" over $X$ and summarizes the overall curvature of the log-likelihood function

## Intuition: Fisher information

::::: columns
::: {.column width="50%"}
**Higher Fisher Information**
```{r}
#| echo: false


ggplot() + 
  geom_function(fun = \(x) log(dnorm(x))) + 
  xlim(c(-3,3)) + 
  labs(
    x = expression(theta),
    y = expression(ln(X,theta))
  )
```
:::

::: {.column width="50%"}
**Lower Fisher Information**

```{r}
#| echo: false


ggplot() + 
  geom_function(fun = \(x) log(dbeta(x, shape1 = 3, shape2=3))) + 
  xlim(c(0,1)) + 
  labs(
    x = expression(theta),
    y = expression(ln(X,theta))
  )
```
:::
:::::

. . . 

Distributions with a "pointier" log-likelihood have higher Fisher information than distributions with lower Fisher information

. . . 

Lower Fisher information $\to$ flat log-likelihood $\to$ lots of estimators "close" to the MLE

## Practice: Fisher information

Find the *Fisher information* for $Y$, where $X \sim Exp(\lambda)$

## Practice: finding CRLB

Let $Y_1, ..., Y_n$ be $n$ Exp($\lambda$) random variables Let $\widehat\lambda = \frac{n}{\sum Y_i}$. How does Var($\widehat\lambda$) compare with the CRLB?


## 

So far, we've found the *bias*, *variance*, and *MSE* **analytically** (with formal mathematical formulas). In practice, we often can't do this. 

. . . 

For example: The median $m$ of an Exponential($\lambda$) distribution satisfies P(X ≤ m) = 0.5. Solving $1 - e^(-\lambda m) = 0.5$ gives $m = ln(2) / \lambda$.

. . . 

This suggests an estimator for $\lambda$ based on the median: $\widehat{\lambda} = \frac{ln(2)}{m}$

. . . 

Finding the analytical variance of $\widehat \lambda$ is complicated. Finding the sampling distribution of $m$ is complicated, and finding the non-linear transformation is also complicated.

## Your turn: simulation 

Use simulation to see whether $\frac{ln(2)}{m}$ (a) is unbiased and (b) achieves the CRLB

