---
title: "18: Classical Hypothesis Tests"
author: "Prof Amanda Luby"
subtitle: "Stat250 S25"
callout-appearance: minimal
knitr:
    opts_chunk: 
      dev: "ragg_png"
      echo: false
      warning: false
      message: false
format:
  pdf:
    include-in-header: 
       - "../preamble.tex"
    toc: false
    number-sections: true
    colorlinks: true
    geometry:
      - top=1in
      - left=1in
      - right=1in
      - bottom=1in
      - heightrounded
    mainfont: "Linux Libertine"
    sansfont: "Linux Libertine"
    mathfont: "Libertinus Math"
    monofont: "Monaco"
fontsize: 11pt
---

```{r}
library(tidyverse)
library(ggformula)
library(knitr)
library(mosaic)
library(patchwork)
library(palmerpenguins)
library(ggridges)
theme_set(theme_minimal())
```

# Logic of Testing

1. Formulate two competing **hypotheses** about the population: the *null hypothesis* and the *alternative hyptohesis*

2. Calculate a **test statistic** that summarizes the relevant information

3. Look at the **behavior of the test statistic** assuming that the null hypothesis is true

4. **Compare** the observed test statistic to the expected behavior (strength of evidence against the null) 

5. State a **conclusion** in context. 


# Permutation Test Recap

**Example:** We are interested in whether there is a difference in the average flipper length between Adelie and Chinstrap penguins. 


```{r}
#| layout-ncol: 2
#| fig-height: 2.5

penguins_subset <- penguins |> filter(species %in% c("Adelie", "Chinstrap")) |> drop_na(flipper_length_mm)

penguins_subset |>
  ggplot(aes(y = species, x = flipper_length_mm, fill = species)) + 
  geom_density_ridges(alpha = .8) + 
  scale_fill_viridis_d(end = .8, option = "plasma") + 
  theme(legend.position = "none")

penguins_subset |>
  group_by(species) |>
  summarize(
    n = n(),
    mean = mean(flipper_length_mm, na.rm = TRUE),
    sd = sd(flipper_length_mm, na.rm = TRUE)
  ) |>
  knitr::kable(digits = 2)

x <- penguins_subset$flipper_length_mm 
observed <- mean(x[penguins_subset$species == "Chinstrap"]) - 
               mean(x[penguins_subset$species == "Adelie"])
```


1. Pool the $n_1 + n_2$ data values

2. Draw a sample of size $n_1$ without replacement, assign those values to Group 1. Assign the remaining $n_2$ values to Group 2. 

3. Calculate the test statistic comparing the samples from the resampled groups.

4. Repeat steps 2 and 3 until we have enough samples.

5. Estimate the p-value as the proportion of times the observed test statistic exceeds the original (observed) test statistic: p-value = $\frac{\text{\# statistics that exceed the original+1}}{\text{\# of statistics in the distribution+1}}$

```{r}
#| echo: true


N <- 10^4 - 1 # Number of permutations to do
sample_size <- nrow(penguins_subset) # Sample size for each permutation (same as data)
x <- penguins_subset$flipper_length_mm # data vector

result <- numeric(N) # Create an empty vector to store results
for (i in 1:N){
  index <- sample(sample_size, 68, replace = FALSE) # Sample indices for group 1
  result[i] <- mean(x[index]) - mean(x[-index]) # Compute differences between groups
}

sum(result >= observed + 1)/(N+1)
```

```{r}
#| fig-width: 3
#| fig-height: 1.5
ggplot() + 
  geom_histogram(aes(x = result), col = "white") + 
  geom_vline(xintercept = observed, 
             linetype = "dashed", 
             col = "darkred")
```

# T-Test for a difference in means

Instead of conducting a permutation test, we can instead **assume** that (a) our data comes from a normal population or (b) that the CLT applies. 

$$\frac{(\bar{X}_1 - \bar{X}_2) - (\mu_1 - \mu_2)}{\sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}} \sim t_{df}$$

**Example:** Use the summary table from p1 to perform a t-test for $H_0: \mu_{\text{Chinstrap}} - \mu_{\text{Adelie}} = 0$ against $H_A: \mu_{\text{Chinstrap}} - \mu_{\text{Adelie}} \ne 0$

\vspace{4in}

```{r}
#| echo: true

t.test(flipper_length_mm ~ species, data = penguins_subset)
```

# T-Test for one mean

**Example:** Do Carls sleep less than 8 hours per night? A sample of Stat 120 students answered the question: "On average, how many hours of sleep do you get on a weeknight?"

```{r}
#| layout-ncol: 2
#| fig-height: 2


survey <- read.csv("https://aloy.rbind.io/data/survey2022.csv")

survey |> 
  ggplot(aes(x = sleep)) + 
  geom_density(fill = "gray") + 
  labs(
    y = ""
  )

survey |>
  summarize(
    n = n(),
    mean = mean(sleep),
    sd = sd(sleep),
    median = median(sleep),
    min = min(sleep),
    max = max(sleep)
  ) |>
  t() |>
  knitr::kable(digits = 2)
```

What is the appropriate **test statistic**? What is the **distribution** of the test statistic under the null hypothesis? 

\vspace{1.5in}

```{r}
#| echo: true 

t.test(~sleep, data = survey, mu = 8, alternative = "less")
```

# Hypothesis tests for binomial data

**Example:** Do Americans support a national health plan? A Kaiser Family Foundation poll for a random sample of US adults in 2019 found that 79% of Democrats, 55% of Independents, and 24% of Republicans supported a generic "National Health Plan." There were 347 Democrats, 298 Republicans, and 617 Independents surveyed. A political pundit on TV claims that a majority of Independents support a National Health Plan. Do these data provide strong evidence to support this type of statement? 

Assume:

Hypotheses:

Test Statistic: 

Null distribution: 

\vspace{1.5in}

```{r}
#| echo: true

binom.test(x = 339, n = 617, p = 0.5, alternative = "greater")
```

CLT: 

Approx null distribution:

Large-sample test statistic:

```{r}
#| echo: true

prop.test(x = 339, n = 617, p = 0.5, alternative = "greater")
``` 

**Is n large enough to use the CLT?**

- Many textbooks suggest _______________ and ________________
- Our textbook suggest _______________ and ________________
- Otherwise, use the exact binomial test

