---
title: "Method of Moments"
subtitle: "Day 07"
title-slide-attributes:
  data-background-gradient: "linear-gradient(to right, #46337e, #440154)"
  data-slide-number: none
author: "Prof Amanda Luby"
format: 
  revealjs:
    incremental: true
editor: 
  markdown: 
    wrap: 72
---

```{r setup}
#| include: false

library(tidyverse)
library(countdown)

library(ggformula)
library(openintro)
library(patchwork)
library(fivethirtyeight)
library(dplyr)
library(tidyr)
library(fontawesome)

library(gridExtra)
library(tidyverse)
library(knitr)
library(mosaic)
library(infer)
library(kableExtra)

library(plotly)

# #440154
knitr::opts_chunk$set(echo = TRUE,
                  message = FALSE,
                  warning = FALSE)

slides_theme = theme_minimal(base_family = "serif", base_size = 16) +
  theme(plot.background = element_rect(fill = "#f0f1eb", colour = NA))
  

theme_set(slides_theme)
```

## Recap

- Observe data $X_1, ..., X_n \sim F_x(x|\theta)$, where $\theta$ is unknown
- Goal: *Estimate* $\theta$ based on the values of $X_i$ by formulating an *estimator* $\hat \theta$
- One technique is to use the *maximum likelihood estimator*, which finds the value of $\theta$ that maximizes the joint probability $\prod_{i=1}^n f_x(x_i | \theta)$
- Today, we'll talk about another technique for estimating $\theta$

# Warm Up

## Which population is more likely?

::::: columns
::: {.column width="50%"}

::: panel-tabset

## Sample

```{r}
#| echo: false

set.seed(0413)

sim_data <- tibble(
  sample = rnorm(40, mean = 20, sd = 5)
) 

sim_data |>
  ggplot(aes(x = sample)) +
  geom_histogram(bins = 15, col = "white") + 
  geom_vline(xintercept = mean(sim_data$sample), lty = "dashed", col = "darkred") + 
  annotate("text", 
           x = mean(sim_data$sample)+.5, 
           y = 12, 
           label = paste("Mean: ", round(mean(sim_data$sample), 2)),
           col = "darkred",
           hjust = 0) + 
  xlim(c(0,40))

sim_data |>
  summarize(
    mean = mean(sample),
    var = var(sample)
  )
```

:::
:::

::: {.column .fragment width="50%"}

::: panel-tabset

## $N(10, 5)$

```{r}
#| echo: false
ggplot() + 
  geom_function(fun = dnorm, args = list(mean = 10, sd = 5)) + 
  xlim(c(0,40))

```

## $N(20, 5)$

```{r}
#| echo: false

ggplot() + 
  geom_function(fun = dnorm, args = list(mean = 20, sd = 5)) + 
  xlim(c(0,40))
```

:::
:::

:::::

## Which population is more likely?

::::: columns
::: {.column width="50%"}

::: panel-tabset

## Sample

```{r}
#| echo: false
sim_data |>
  ggplot(aes(x = sample)) +
  geom_histogram(bins = 15, col = "white") + 
  geom_vline(xintercept = mean(sim_data$sample), lty = "dashed", col = "darkred") + 
  annotate("text", 
           x = mean(sim_data$sample)+.5, 
           y = 12, 
           label = paste("Mean: ", round(mean(sim_data$sample), 2)),
           col = "darkred",
           hjust = 0) + 
  xlim(c(0,40))

sim_data |>
  summarize(
    mean = mean(sample),
    var = var(sample)
  )
```

:::
:::

::: {.column .fragment width="50%"}

::: panel-tabset

## $N(20, 1.5^2)$

```{r}
#| echo: false
ggplot() + 
  geom_function(fun = dnorm, args = list(mean = 20, sd = 1.5^2)) + 
  xlim(c(-10,50))

```

## $N(20, 4^2)$

```{r}
#| echo: false

ggplot() + 
  geom_function(fun = dnorm, args = list(mean = 20, sd = 4^2)) + 
  xlim(c(-10,50))
```

:::
:::

:::::

## Method of Moments (MoM) 

- Mean of sample data should match mean of theoretical distribution
- Variance of sample data should match variance of theoretical distribution
- Skewness of sample data should match skewness of theoretical distribution
- $\vdots$
- (for as many unknown parameters as we have)


