---
title: "06: More on MLEs"
author: "Prof Amanda Luby"
subtitle: "Stat250 S25"
callout-appearance: minimal
knitr:
    opts_chunk: 
      dev: "ragg_png"
      echo: false
      warning: false
      message: false
format:
  pdf:
    include-in-header: 
       - "../preamble.tex"
    toc: false
    number-sections: true
    colorlinks: true
    geometry:
      - top=1in
      - left=1in
      - right=1in
      - bottom=1in
      - heightrounded
fontfamily: libertine
fontsize: 11pt
---


```{r, echo = FALSE}
library(tidyverse)
library(Sleuth3)
library(mosaic)
library(infer)
```

# More on the likelihood function and MLEs

**Example:** Recall that the likelihood function for $n$ iid Bernoulli($\theta$) random variable is $L(\theta) = \theta^{\sum x_i} (1-\theta)^{n - \sum x_i}$

Scenario 1: 0 "Yes" responses
Scenario 2: 6 "Yes" responses 

```{r}
#| fig-width: 3
#| fig-height: 2
#| fig-align: 'left'
ggplot() + 
  xlim(c(0,1)) + 
  ylim(c(0,1)) + 
  theme_bw()
```


**Exercise** Let $X_1, ..., X_n$ be an iid random sample from a distribution with PDF 

$$f(x|\theta)= (\theta + 1)x^\theta, 0 \le x \le 1$$

$$L(\theta) =$$
$$\hat\theta_{MLE}=$$

Suppose we observe a sample of size 5: {.83, .49, .72, .57, .66}. Find the maximum likelihood estimate and verify with a graph or numerical approximation

\vspace{3in}

# Uniform distribution

Find the MLE for $Y_1, ..., Y_n \sim \text{Unif}(0, \theta)$

\vspace{5in}

**Exercise:** Is the likelihood function a probability distribution? Why or why not?

\vspace{2in}

# Finding the MLE when more than one parameter is unknown

If the pdf or pmf that we're using has two or more parameters, say $\theta_1$ and $\theta_2$, finding MLEs for the $\theta_i$'s requires the solution of a sest of simultaneous equations. We would typically need to solve the following system: 

$$\frac{\partial}{\partial \theta_1} \ln L(\theta_1, \theta_2) = 0$$


$$\frac{\partial}{\partial \theta_2} \ln L(\theta_1, \theta_2) = 0$$

**Example:** Suppose a random sample of size $n$ is drawn from the two parameter normal pdf

$$f_y(y|\mu, \sigma^2) = \frac{1}{\sqrt{2\pi \sigma}} \exp(\frac{-(y-\mu)^2}{2\sigma^2})$$

find the MLEs $\hat{\mu}$ and $\hat\sigma^2$
