---
title: "02: Permutation Tests"
subtitle: "Reference: MSwRR Ch 3"
date: "Stat250-S25"
author: "Prof Amanda Luby"
callout-appearance: minimal
knitr:
    opts_chunk: 
      dev: "ragg_png"
      echo: false
      warning: false
      message: false
format:
  pdf:
    include-in-header: 
       - "../preamble.tex"
    toc: false
    number-sections: true
    colorlinks: true
    geometry:
      - top=1in
      - left=1in
      - right=1in
      - bottom=1in
      - heightrounded
fontfamily: libertine
fontsize: 11pt
---


```{r, echo = FALSE}
library(tidyverse)
library(Sleuth3)
library(mosaic)
library(infer)
```

# Example: Motivation Experiment

```{r poem_boxes, fig.height=2.5, fig.width=4, echo=FALSE, fig.align='left', out.width = "50%"}
orig.means <- case0101 %>%
  group_by(Treatment) %>%
  summarize(mean = mean(Score))
ggplot(data = case0101) +
  geom_boxplot(mapping = aes(x = Treatment, y = Score, fill = Treatment), alpha = 0.5) +
  geom_point(data = orig.means, mapping = aes(x = Treatment, y = mean)) +
  scale_fill_viridis_d() +
  theme_minimal() + 
  labs(caption = "Dot denotes mean") +
  coord_flip() +
  theme(legend.position = "none")
```

```{r echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
knitr::kable(favstats(Score ~ Treatment, data = case0101), format = "markdown", digits = 1)
```

# Logic of Hypothesis Testing

1. Formulate two competing _________________ about the _________________

2. Calculate a __________________________ summarizing the relevant information to the claims

3. Look at the ______________________________________ assuming that the initial claim is true

4. ______________________ the observed test statistic to the expected behavior (the distribution created in step 3). 


**Example:** Write down the competing hypotheses being tested by the researchers, both in words and using proper notation.

\vspace{2in}

::: callout-note
## P-value

\vspace{1in}
:::

```{r}
#| echo: false
#| fig-width: 4
#| fig-height: 3
#| fig-align: 'left'

perm_dsn <-   case0101 %>%
  specify(Score ~ Treatment) %>% # 
  hypothesize(null = "independence") %>%
  generate(reps = 1999, type = "permute") %>%
  calculate(stat = "diff in means", order = c("Intrinsic", "Extrinsic"))

null_dsn <- perm_dsn %>%
  visualize(bins = 40) +
  scale_x_continuous(breaks = -5:5) +
  labs(title = "Distribution of difference in means",
       subtitle = "Assuming equal means, based on 1,999 trials",
       x = "Test statistics") +
  theme_minimal(base_size = 10) + 
  theme(panel.grid.minor.x = element_blank(),
        panel.grid.major.x = element_blank())
null_dsn
```

# R Code

Use this space to jot down any notes about the R code we talked through
